Think deep for reef resilience

### 2025-09-13 paper/paper_draft.qmd
Standardize timeseries -> time series
Sometimes timeseries was used, other times it is "time series". The latter was more common so I standardized everything to "time series".

ADRIA-CoralBlox was run for 4096 scenarios for each GCM (20,480 total model runs). Here, a
scenario represents a unique combination of assumed environmental (namely one of the five
DHW trajectories) and ecological factors including coral growth and mortality rates, and
coral thermal tolerances. Scenarios were generated using the low-discrepancy Sobol' sampling
method [@sobol1967; @sobol1993] with Owen Scrambling [@owen1995], a randomised Quasi Monte
Carlo approach which balances scenario exploration with improved convergence properties
compared to pseudo-random sampling. A detailed comparative analysis of sampling methods can
be found in [@kucherenko2015]. For the purpose of scenario assessment, the resulting coral
cover time series for each model run were scaled by the total area of each reef to represent
cover as a percentage of the total estimated reef area. The scenario-median timeseries for
each reef was used for further analyses. These time series of coral cover were then
clustered based on their temporal behaviour.

@@

To assess the impact of reef connectivity on modelled outcomes we summarise incoming
connectivity (weighted incoming connectivity), outgoing connectivity (outgoing connection
strength), the ratio of source-to-sink connectivity, and larval retention probability for
each reef. Weighted incoming connectivity is calculated by summing the product of each
connection strength incoming to a target reef and the mean timeseries absolute coral cover
(km^2^) of each matching source reef. This metric provides an approximation of overall
larval flow into a reef, with high values indicating the target reef is supplied high
volumes of larvae via strong connections and low values indicating the target reef has weak
(or no) incoming connections or is supplied by small source populations. A limitation of
weighted incoming connectivity is it does not incorporate the proportion of the source
reef's population that is reproducing and producing larvae, instead taking the total coral
cover area as an indicator of larval productivity. Outgoing connection strength is
calculated as the sum of all outgoing transition probabilities from a target reef. Outgoing
connection strength has a theoretical maximum of one (all larvae reach settlement on sink
reefs), however the large size of the GBR combined with larval mortality considerations
result in outgoing connection strength values much lower than one. The ratio of
source-to-sink connectivity ($O:I$) for a target reef represents the strength of the reef's
outgoing larval flow relative to it's incoming larval flow. This is calculated as:
$O:I = \frac{S_O/N_O}{S_I/N_I}$. Where $S$ represents the sum of connectivity probabilities,
$N$ represents the number of connections, $O$ denotes outgoing connectivity from the target
reef and $I$ denotes incoming connectivity to the target reef. If a reef has no incoming
connections, the $O:I$ equals the $S_O/N_O$ value. Incoming connections in this calculation,
do not directly represent the larval supply gained as the breeding populations from which
connections are coming from are not represented. A source-to-sink ratio greater than 1 (a
value greater than 0 on a log scale) indicates a reef has greater outgoing than incoming
larval flow. Reef larval retention probability indicates the proportion of spawned larvae
from a target reef that remain to settlement.

@@

| Metric | Used in | Definition | Justification |
|-----------------------|-----------------------|-----------------------|-----------------------|
| Percentage coral cover time series | Timeseries clustering, Carbonate budget analysis | Total coral cover as a percentage of reef area | To summarise coral population health over time |
| Median depth | Time series cluster assessment Carbonate budget analysis | Reef median depth as calculated with zonal raster analysis | Depth is a critical form of environmental refugia from high thermal stress |
| Mean time series DHW | Time series cluster assessment | Reef mean DHW value across time series | Resilient reefs may experience reduced surface DHWs |
| Source - to - sink connectivity ratio | Time series cluster assessment | Ratio of reef outgoing to incoming connectivity | Reefs with greater source than sink connectivity may have lower recovery |
| Weighted incoming connectivity | Time series cluster assessment, Carbonate budget analysis | Incoming connectivity strength weighted by mean time series coral cover of source reefs | High incoming larval flow assists reef recovery |
| Weighted outgoing connectivity | Time series cluster assessment | Outgoing connectivity strength weighted by mean coral cover | High outgoing larval flow may lower recovery capacity of the target reef |
| Carrying capacity | Time series cluster assessment | Reef habitable area \[m^2^\] | Captures the size of the viable reef extent |

@@

The CID timeseries clustering metric corrects the Euclidean distance between pairs of time
series (which accounts for magnitude differences) with the complexity of each time series
(obtained through a measure of their variability) to ensure reefs are clustered based on
their underlying temporal dynamics rather than simply their average coral cover. The process
creates a distance matrix of n \* n dimensions (where n is the number of time series in the
group) that is clustered using a k-medoids clustering algorithm [cf. @steinmann2020]. The
three reef clusters are labelled with qualitative-ordered clusters from Low, Medium and High
based on their median cover relative to each other cluster over the 2030-2060 period. The
selected time frame prevents the classification from being overly influenced by any steep
declines in the initial years of simulations and avoids the post-2060 time period where
reefs are projected to be under severe stress and in a declined state [@bozec2025].

>>>>

ADRIA-CoralBlox was run for 4096 scenarios for each GCM (20,480 total model runs). Here, a
scenario represents a unique combination of assumed environmental (namely one of the five
DHW trajectories) and ecological factors including coral growth and mortality rates, and
coral thermal tolerances. Scenarios were generated using the low-discrepancy Sobol' sampling
method [@sobol1967; @sobol1993] with Owen Scrambling [@owen1995], a randomised Quasi Monte
Carlo approach which balances scenario exploration with improved convergence properties
compared to pseudo-random sampling. A detailed comparative analysis of sampling methods can
be found in [@kucherenko2015]. For the purpose of scenario assessment, the resulting coral
cover time series for each model run were scaled by the total area of each reef to represent
cover as a percentage of the total estimated reef area. The scenario-median time series for
each reef was used for further analyses. These time series of coral cover were then
clustered based on their temporal behaviour.

@@

To assess the impact of reef connectivity on modelled outcomes we summarise incoming
connectivity (weighted incoming connectivity), outgoing connectivity (outgoing connection
strength), the ratio of source-to-sink connectivity, and larval retention probability for
each reef. Weighted incoming connectivity is calculated by summing the product of each
connection strength incoming to a target reef and the mean time series absolute coral cover
(km^2^) of each matching source reef. This metric provides an approximation of overall
larval flow into a reef, with high values indicating the target reef is supplied high
volumes of larvae via strong connections and low values indicating the target reef has weak
(or no) incoming connections or is supplied by small source populations. A limitation of
weighted incoming connectivity is it does not incorporate the proportion of the source
reef's population that is reproducing and producing larvae, instead taking the total coral
cover area as an indicator of larval productivity. Outgoing connection strength is
calculated as the sum of all outgoing transition probabilities from a target reef. Outgoing
connection strength has a theoretical maximum of one (all larvae reach settlement on sink
reefs), however the large size of the GBR combined with larval mortality considerations
result in outgoing connection strength values much lower than one. The ratio of
source-to-sink connectivity ($O:I$) for a target reef represents the strength of the reef's
outgoing larval flow relative to it's incoming larval flow. This is calculated as:
$O:I = \frac{S_O/N_O}{S_I/N_I}$. Where $S$ represents the sum of connectivity probabilities,
$N$ represents the number of connections, $O$ denotes outgoing connectivity from the target
reef and $I$ denotes incoming connectivity to the target reef. If a reef has no incoming
connections, the $O:I$ equals the $S_O/N_O$ value. Incoming connections in this calculation,
do not directly represent the larval supply gained as the breeding populations from which
connections are coming from are not represented. A source-to-sink ratio greater than 1 (a
value greater than 0 on a log scale) indicates a reef has greater outgoing than incoming
larval flow. Reef larval retention probability indicates the proportion of spawned larvae
from a target reef that remain to settlement.

@@

| Metric | Used in | Definition | Justification |
|-----------------------|-----------------------|-----------------------|-----------------------|
| Percentage coral cover time series | Time series clustering, Carbonate budget analysis | Total coral cover as a percentage of reef area | To summarise coral population health over time |
| Median depth | Time series cluster assessment Carbonate budget analysis | Reef median depth as calculated with zonal raster analysis | Depth is a critical form of environmental refugia from high thermal stress |
| Mean time series DHW | Time series cluster assessment | Reef mean DHW value across time series | Resilient reefs may experience reduced surface DHWs |
| Source - to - sink connectivity ratio | Time series cluster assessment | Ratio of reef outgoing to incoming connectivity | Reefs with greater source than sink connectivity may have lower recovery |
| Weighted incoming connectivity | Time series cluster assessment, Carbonate budget analysis | Incoming connectivity strength weighted by mean time series coral cover of source reefs | High incoming larval flow assists reef recovery |
| Weighted outgoing connectivity | Time series cluster assessment | Outgoing connectivity strength weighted by mean coral cover | High outgoing larval flow may lower recovery capacity of the target reef |
| Carrying capacity | Time series cluster assessment | Reef habitable area \[m^2^\] | Captures the size of the viable reef extent |

@@

The CID time series clustering metric corrects the Euclidean distance between pairs of time
series (which accounts for magnitude differences) with the complexity of each time series
(obtained through a measure of their variability) to ensure reefs are clustered based on
their underlying temporal dynamics rather than simply their average coral cover. The process
creates a distance matrix of n \* n dimensions (where n is the number of time series in the
group) that is clustered using a k-medoids clustering algorithm [cf. @steinmann2020]. The
three reef clusters are labelled with qualitative-ordered clusters from Low, Medium and High
based on their median cover relative to each other cluster over the 2030-2060 period. The
selected time frame prevents the classification from being overly influenced by any steep
declines in the initial years of simulations and avoids the post-2060 time period where
reefs are projected to be under severe stress and in a declined state [@bozec2025].
